Sea: A lightweight data-placement library for Big Data scientific computing

### 2022-05-06 paper/sea-comp/paper.tex
additional fixes

% The paper headers
\markboth{Journal of \LaTeX\ Class Files,~Vol.~14, No.~8, August~2015}%
{Hayot-Sasson\MakeLowercase{\textit{et al.}}: Bare Demo of IEEEtran.cls for
Computer Society Journals} \IEEEtitleabstractindextext{%
\begin{abstract}
    The recent influx of scientific data has lead to scientific computing
transitioning from compute intensive to data intensive. Whereas many Big Data
Frameworks exist that minimize the cost of data transfers, few scientific
applications adopt data-placement strategies as the efforts required to update
the existing applications is significant.
We developed Sea as a means to enable data-placement strategies for scientific
applications executing on HPC clusters without the need to reinstrument
workflows. Sea leverages GNU C library interception to intercept POSIX-compliant
file system calls made by the applications. We designed a performance model and
evaluated the performance of Sea on a synthetic data-intensive application
processing a representative neuroimaging dataset (the Big Brain). Our results
demonstrate that Sea significantly improves performance, up to a factor of
3$\times$.

@@

\IEEEraisesectionheading{\section{Introduction}\label{sec:introduction}}
\IEEEPARstart{E}{fficient} data-placement strategies have become essential in
minimizing Big Data processing overheads. These strategies, such as data
locality and in-memory computing, improve application runtime by ensuring data
accesses occur on the most efficient available storage. Existing Big Data
frameworks, such as MapReduce~\cite{dean2008mapreduce}, Apache
Spark~\cite{zaharia2016apache} and Dask~\cite{rocklin2015dask}, improve runtime
performance through the incorporation of these data-placement strategies.

@@

Our goal is to provide a data-placement solution to standardized scientific
computing applications executing on High Performance Computing (HPC) clusters.
The tools of particular interest to us consist of various tasks that communicate
to each other via a POSIX-compliant file system. Whereas the tools can be
platform-agnostic, we assume users will leverage them within an HPC context,
particularly when processing large volumes of data, due to their inherent
reliance on POSIX-compliance.

We developed Sea, a user-space data-placement library that provides efficient
data placement without reinstrumentation of the underlying application. Given a
predefined storage hierarchy, Sea can prefetch, cache, flush, and evict data to
and from a permanent storage location. To provide a
frame-of-reference for Sea's performance on HPC systems, we designed a
performance model. We then evaluated Sea's performance on a synthetic Big Data
application running on a representative neuroimaging dataset, comparing its
performance with the bounds delineated by the model.

\section{Related Work}
\subsection{HPC Infrastructure}
      The general structure of HPC clusters complicates the deployment of Big
      Data frameworks. Typical HPC clusters consist of distinct storage and
      compute nodes. While compute nodes may also have local storage, there is
      no distributed file system like the Hadoop Distributed File System
      (HDFS)~\cite{shvachko2010hadoop} or Alluxio~\cite{alluxio} to facilitate
      data locality. Furthermore, access to these compute nodes is temporary,
      with allocation duration enforced by a batch scheduler. Therefore, any
      data written to compute nodes are rendered inaccessible after the
      allocation is terminated.

      The storage layer found on HPC clusters consists of a high-performance
      parallel file system (PFS) (e.g. Lustre~\cite{lustre}). These file systems
      are shared amongst all compute nodes within a cluster, typically connected
      to each node via high-performance network interconnect such as
      InfiniBand~\cite{infiniband}. I/O overheads of shared PFS can be incurred
      in many places, from the shared network to the bandwidth and latency of
      the storage devices.  With Lustre specifically, there typically are many
      data nodes, known as Object Storage Servers (OSS) which contain several
      storage devices, known as Object Storage Targets (OST). All file metadata
      is maintained within a separate node known as the Metadata Server (MDS)
      and stored within a device referred to as the Metadata Target (MDT).
      Although data transfers can be communicated directly to the corresponding
      OST, the clients need to first communicate with the MDS to determine which
      OST to communicate with. This can result in major data transfer overheads,
      particularly when making numerous requests to the metadata server.

@@

      Strategies to mitigate the impacts of data transfers in data-intensive
      workloads are required to ensure optimal processing time. Early Big Data
      frameworks, notably MapReduce~\cite{dean2008mapreduce} and Apache
      Spark~\cite{zaharia2016apache} have implemented and popularized
      data-management strategies to minimize data transfers at processing time.
      These strategies are known as data locality and in-memory computing.

      Data locality ensures that compute tasks are scheduled nearest to where
      the data is located. Historically, storage and compute nodes were distinct
      layers, transferring data to the compute location whenever necessary. Rather than transferring
      large amounts of data over the network to compute tasks, which could incur
      significant overheads, Big Data frameworks ensure that data is stored
      directly on the compute nodes. When a compute task requires access to
      specific data, the scheduler sends the task to the nearest available node
      to the data, thereby minimizing any cost of network-related data
      transfers. This strategy is not only used by Big Data Frameworks such as
      Hadoop MapReduce, Apache Spark, and Dask~\cite{rocklin2015dask}, but also
      enabled by file systems such as HDFS and Alluxio.

@@

      While Big Data frameworks and filesystems are not commonly used in HPC,
      HPC clusters are typically Linux-based and benefit from the Linux page
      cache, which can provide data locality and in-memory computing at a
      limited capacity. Furthermore, HPC architectures may also provide an
      intermediate storage layer, known as a Burst Buffer, to improve
      application I/O-related overheads. To facilitate an applications
      interaction between application and storage layer, distributed file
      systems have been developed.

\subsection{The Linux page cache}
 While scientific applications do not customarily leverage data-placement
      strategies and may perform large over-the-network data transfers, they may
      still benefit from in-memory computing and data locality through the Linux
      Page Cache~\cite{pagecache}. Similarly to other file systems, Lustre
      leverages the compute node page cache to reduce I/O overheads. System
      memory is composed of two components: 1) anonymous memory and 2) page
      cache. Anonymous memory consists of all application-related objects,
      whereas page cache consists of recently accessed file data. When a file is
      read, that file is loaded up into the page cache to be flagged for
      eviction based on a least recently used (LRU) policy. That means
      subsequent accesses to that data may be done entirely in memory so long as
      that data has not already been evicted. Similarly, for writes to a file
      system with writeback cache enabled, the file will be written to memory
      completing the write operation once the file has been written entirely to
      memory. That file will then be flushed asynchronously to the appropriate
      storage device. As system memory may get overloaded with too many write
      requests, there is a limit to the amount of written data that can exist in
      memory, known as the dirty\_ratio. Furthermore, applications producing too
      many write requests may be throttled by the system.

@@

      However, there exist Big Data file systems, such as
      XtreemFS~\cite{xtreemfs}, that exist entirely within user space, are
      POSIX-compliant, and can use alternative methods to FUSE to function. For
      instance, XtreemFS uses the LD\_PRELOAD trick to intercept file system
      calls made to the GNU C library (glibc). There are limitations to using
      glibc interception; the application must be dynamically linked and make
      glibc calls. However, statically-linked applications are uncommon and
      applications interacting with POSIX-compliant file systems on Linux
      machines, the predominant OS of HPC clusters, will make glibc calls.
      Nevertheless, there are alternatives to glibc intercept that can bypass
      the aforementioned issues, such as system call interception, but they
      result in greater overheads~\cite{quinson}.

@@

      BurstFS~\cite{burstfs} and GekkoFS~\cite{gekkofs} are two user-space file
      systems that are specifically designed for Burst Buffers. That is, both of
      these file systems exist only during the time-window in which the
      resources are allocated. Furthermore, they both exist entirely within
      user-space using glibc interception. While these implementations simplify
      usage of burst buffers for users, their installation and configuration is
      similarly out-of-scope for a typical HPC user.

@@

Sea requires minimal configuration for use. In order to work, Sea requires the
user to specify at least two storage devices, a fast temporary one (i.e. cache)
and a slower long-term storage one. This could be RAM and SSD if working on a
single node, or a compute-local SSD and a shared parallel file system, in the
case of an HPC cluster. Ideally, a user will provide a multitude of ephemeral
storage devices to improve Sea's efficacy. Furthermore, to maximize usage of the
fastest available storage devices, Sea will allow the user to outline which
files can be removed from short-term storage in addition to which files need to
be materialized onto long-term storage. Eviction of files is important as this
will allow Sea to maximize usage of short-term storage.

@@

      \end{tabular}
      \caption{Sea's memory management modes}
      \label{table:sea-comp:fe}
      \end{table}
%\todo{add diagrams here?}
Memory-management is an integral component of Sea as deallocating data from cache
makes room for future files to be stored within the higher-performance storage devices.
Generalizing an efficient memory-management strategy across all applications is a challenging task.
Since Sea works on a per-application basis, memory-management in Sea is application-specific, configured
via two files: \texttt{.sea\_flushlist} and \texttt{.sea\_evictlist}.

Memory management in Sea consists of four different modes, as described in Table~\ref{table:sea-comp:fe}: flush-only,
evict-only, flush-and-evict and keep. A flush-only is synonymous with a file
copy in that a file that already exists within a Sea cache will be copied to
persistent storage. This operation is particularly useful when a file will be
reused by the application, and thus benefits from remaining in cache, but also
needs to be communicated between other nodes or needs to be persisted for
post-processing analysis. To flush-only a file, the file must only be specified
in the \texttt{.sea\_flushlist}.

Evict-only is equivalent to a remove operation. This mode is for files that are
located within a Sea cache, but do not need to be persisted and will not be
reused by the pipeline and therefore do not need to occupy cache space.
Evict-only is useful for removing unnecessary application log files. To
evict-only a file, the file must only be specified in the
\texttt{.sea\_evictlist}.

Similar to the flush-only, the flush-and-evict mode copies data from cache to
persistent storage. However, the assumption here is that these files will not be
reused by the application and therefore need not occupy space in cache.
Flush-and-evict is therefore synonymous with a move operation and is invoked on
files that are specified in both the \texttt{.sea\_flushlist} and
\texttt{.sea\_evictlist}.

@@

      Here, $M_{n}$ represents the makespans of the different possible storage
      levels (e.g. tmpfs, NVMe, SSD, HDD, Lustre) and $M_{S}$ represent the
      makespan of an application reading and writing to Sea. For our model, we
      will assume 3 storage layers: 1) fast tmpfs, 2) intermediate local SSD
      storage, and 3) slow parallel file system layer. It is important to note that
      the model assumes that I/O never occurs in distinct storage layers in parallel.
      Although Sea tries to exhaust the layers, starting from the fastest, it is nevertheless
      possible that Sea writes to two storage layers at the same time, for instance, when one becomes full
      and another parallel thread needs to perform a write.

>>>>

% The paper headers
\markboth{Journal of \LaTeX\ Class Files,~Vol.~14, No.~8, August~2015}%
{Hayot-Sasson\MakeLowercase{\textit{et al.}}: Bare Demo of IEEEtran.cls for
Computer Society Journals} \IEEEtitleabstractindextext{%
\begin{abstract}
    The recent influx of open scientific data has contributed to the
    transitioning of scientific computing from compute intensive to data intensive.
    Whereas many Big Data
frameworks exist that minimize the cost of data transfers, few scientific
applications integrate these frameworks or adopt data-placement strategies to mitigate the costs.
Scientific applications commonly rely on well-established command-line tools that would require complete reinstrumentation
in order to incorporate existing frameworks.
We developed Sea as a means to enable data-placement strategies for scientific
applications executing on HPC clusters without the need to reinstrument
workflows. Sea leverages GNU C library interception to intercept POSIX-compliant
file system calls made by the applications. We designed a performance model and
evaluated the performance of Sea on a synthetic data-intensive application
processing a representative neuroimaging dataset (the Big Brain). Our results
demonstrate that Sea significantly improves performance, up to a factor of
3$\times$.

@@

\IEEEraisesectionheading{\section{Introduction}\label{sec:introduction}}
\IEEEPARstart{E}{fficient} data-placement strategies have become essential to
minimizing Big Data processing overheads. These strategies, such as data
locality and in-memory computing, improve application runtime by ensuring data
accesses occur on the most efficient available storage. Existing Big Data
frameworks, such as MapReduce~\cite{dean2008mapreduce}, Apache
Spark~\cite{zaharia2016apache} and Dask~\cite{rocklin2015dask}, improve runtime
performance through the incorporation of data-placement strategies.

@@

Our goal is to provide a data-placement solution to standardized scientific
computing applications executing on High Performance Computing (HPC) clusters.
The applications of interest to us consist of various tasks that communicate
to each other via a POSIX-compliant file system. Whereas the application can be
platform-agnostic, we assume users will leverage them within an linux-based HPC context,
particularly when processing large volumes of data, due to their inherent
reliance on POSIX-compliance.

We developed Sea, a user-space data-placement library that provides efficient
data placement without reinstrumentation of the underlying application. Given a
predefined storage hierarchy, Sea can prefetch, cache, flush, and evict data to
and from a permanent storage location.

To provide a
frame-of-reference for Sea's performance on HPC systems, we designed a
performance model. We then evaluated Sea's performance on a synthetic Big Data
application running on a representative neuroimaging dataset, comparing its
performance with the bounds delineated by the model.

\section{Related Work}
\subsection{HPC Infrastructure}
      The general structure of HPC clusters complicates the deployment of Big
      Data frameworks. Typical HPC clusters consist of distinct storage and
      compute nodes. While compute nodes may also have local storage, there is
      no distributed file system like the Hadoop Distributed File System
      (HDFS)~\cite{shvachko2010hadoop} or Alluxio~\cite{alluxio} to facilitate
      data locality. Furthermore, access to these compute nodes is ephemeral,
      with allocation duration enforced by a batch scheduler. Therefore, any
      data written to compute nodes are rendered inaccessible after the
      allocation is terminated.

      The storage layer found on HPC clusters consists of a high-performance
      parallel file system (PFS) (e.g. Lustre~\cite{lustre}). These file systems
      are shared amongst all compute nodes within a cluster, typically connected
      to the nodes via high-performance network interconnect like
      InfiniBand~\cite{infiniband}. I/O overheads of shared PFS can be incurred
      in many places, from the shared network to the bandwidth and latency of
      the storage devices.  With Lustre specifically, there typically are many
      data nodes, known as Object Storage Servers (OSS) which contain several
      storage devices, known as Object Storage Targets (OST). All file metadata
      is maintained within a separate node known as the Metadata Server (MDS)
      and stored within a device referred to as the Metadata Target (MDT).
      Although data transfers can be communicated directly to the corresponding
      OST, the clients need to first communicate with the MDS to determine which
      OST to communicate with. This can result in major data transfer overheads,
      particularly when making numerous requests to the metadata server.

@@

Early Big Data
      frameworks, notably MapReduce~\cite{dean2008mapreduce} and Apache
      Spark~\cite{zaharia2016apache} have implemented and popularized
      data-management strategies to minimize data transfers at processing time.
      These strategies are known as data locality and in-memory computing.
      Data locality ensures that compute tasks are scheduled nearest to where
      the data is located. Historically, storage and compute nodes were distinct
      layers, transferring data to the compute location whenever necessary. Rather than transferring
      large amounts of data over the network to compute tasks, which could incur
      significant overheads, Big Data frameworks ensure that data is stored
      directly on the compute nodes. When a compute task requires access to
      specific data, the scheduler sends the task to the nearest available node
      to the data, thereby minimizing any cost of network-related data
      transfers. This strategy is not only used by Big Data Frameworks such as
      Hadoop MapReduce, Apache Spark, and Dask~\cite{rocklin2015dask}, but also
      enabled by file systems such as HDFS and Alluxio.

@@

      While Big Data frameworks and filesystems are not commonly used in HPC,
      HPC clusters are typically Linux-based and benefit from the Linux page
      cache, which can provide data locality and in-memory computing at a
      limited capacity. Furthermore, HPC architectures may also provide an
      intermediate storage layer, known as a Burst Buffer, to improve
      application I/O-related overheads. To facilitate an application's
      interaction between application and storage layer, distributed file
      systems have been developed.

\subsection{The Linux page cache}
 While scientific applications do not customarily leverage data-placement
      strategies and may perform large over-the-network data transfers, they may
      still benefit from in-memory computing and data locality through the Linux
      Page Cache~\cite{pagecache}. Similarly to other file systems, Lustre
      leverages the compute node page cache to reduce I/O overheads. System
      memory is composed of two components: 1) anonymous memory and 2) page
      cache. Anonymous memory consists of all application-related objects,
      whereas page cache consists of recently accessed file data. When a file is
      read, that file is loaded up into the page cache to be flagged for
      eviction based on a least recently used (LRU) policy. That means
      subsequent accesses to that data may be done entirely in memory so long as
      the data has not already been evicted. Similarly, for writes to a file
      system with writeback cache enabled, the file will be written to memory
      completing the write operation once the file has been written entirely to
      memory. That file will then be flushed asynchronously to the appropriate
      storage device. As system memory may get overloaded with too many write
      requests, there is a limit to the amount of written data that can exist in
      memory, known as the dirty\_ratio. Furthermore, applications producing too
      many write requests may be throttled by the system.

@@

      However, there exist Big Data file systems, such as
      XtreemFS~\cite{xtreemfs}, that exist entirely within user space, are
      POSIX-compliant, and can use alternative methods to FUSE to function. For
      instance, XtreemFS uses the LD\_PRELOAD trick to intercept file system
      calls made to the GNU C library (glibc). There are limitations to using
      glibc interception; the application must be dynamically-linked and make
      glibc calls. However, statically-linked applications are uncommon and
      applications interacting with POSIX-compliant file systems on Linux
      machines, the predominant OS of HPC clusters, will make glibc calls.
      Nevertheless, alternatives to glibc intercept exist that can bypass
      the aforementioned issues, such as system call interception, but they
      result in greater overheads~\cite{quinson}.

@@

      BurstFS~\cite{burstfs} and GekkoFS~\cite{gekkofs} are two user-space file
      systems that are specifically designed with HPC Burst Buffers. That is, both of
      these file systems are designed to support the ephemeral nature of HPC compute nodes and their associated storage.
      Additionally, these implementations exist entirely in user-space leveraging the LD\_PRELOAD trick to intercept
      necessary library calls. Whereas these libraries incorporate a client-server architecture to aid in transferring data
      between nodes, Sea opts for a more lightweight approach ensuring decentralization, statelessness and leveraging underlying
      file systems for communication and consistency.

@@

Sea requires minimal configuration for use. In order to work, Sea requires the
user to specify at least two storage devices, a fast temporary device used as cache
and a slower long-term storage device. This could be RAM and SSD if working on a
single node, or a compute-local SSD and a shared PFS, in the
case of an HPC cluster. Ideally, a user will provide a multitude of ephemeral
storage devices to improve Sea's efficacy. Furthermore, to maximize usage of the
fastest available storage devices, Sea will allow the user to outline which
files can be removed from short-term storage in addition to which files need to
be materialized onto long-term storage. Eviction of files is important as this
will allow Sea to maximize usage of short-term storage.

@@

      \end{tabular}
      \caption{Sea's memory management modes}
      \label{table:sea-comp:fe}
      \end{table}
%\todo{add diagrams here?}
Memory-management is an integral component of Sea as de-allocating data from cache
makes room for future files to be stored within the higher-performance storage devices.
Generalizing an efficient memory-management strategy across all applications is a challenging task.
Since Sea works on a per-application basis, memory-management in Sea is application-specific, configured
via two files: \texttt{.sea\_flushlist} and \texttt{.sea\_evictlist}.

Memory management in Sea consists of four different modes, as described in Table~\ref{table:sea-comp:fe}: copy,
remove, move and keep. A copy copies a file located within the Sea cache to
persistent storage. This operation is particularly useful when a file will be
reused by the application, and thus benefits from remaining in cache, but also
needs to be communicated between other nodes or needs to be persisted for
post-processing analysis. To copy a file, the file must only be specified
in the \texttt{.sea\_flushlist}.

Remove is for files that are
located within a Sea cache, but do not need to be persisted and will not be
reused by the pipeline and therefore do not need to occupy cache space.
Remove is useful for removing unnecessary application log files. To
remove a file, the file must only be specified in the
\texttt{.sea\_evictlist}.

Similar to the copy, the move mode copies data from cache to
persistent storage. However, the assumption here is that these files will not be
reused by the application and therefore need not occupy space in cache.
Move is therefore synonymous with a copy-and-remove operation and is invoked on
files that are specified in both the \texttt{.sea\_flushlist} and
\texttt{.sea\_evictlist}.

@@

      Here, $M_{n}$ represents the makespans of the different possible storage
      levels (e.g. tmpfs, NVMe, SSD, HDD, Lustre) and $M_{S}$ represent the
      makespan of an application reading and writing to Sea. For our model, we
      will assume three storage layers: 1) fast tmpfs, 2) intermediate local SSD
      storage, and 3) slow parallel file system layer. It is important to note that
      the model assumes that I/O never occurs in distinct storage layers in parallel.
      Although Sea tries to exhaust the layers, starting from the fastest, it is nevertheless
      possible that Sea writes to two storage layers at the same time, for instance, when one becomes full
      and another parallel thread needs to perform a write.
